---
title: "Confirmatory Data Analysis"
Author: "Virinchi Doddapaneni"
date: "March 15, 2026"
date-modified: "last-modified"
format: 
  html:
    theme: cosmo  
---

# Confirmatory Data Analysis

## 1. Installing and launching required R packages

Before performing the confirmatory analysis, the required R packages are loaded. These packages support data manipulation (tidyverse), statistical visualization (ggplot2), interactive graphics (plotly), scaling functions (scales), and correlation visualization (corrplot).

```{r}
pacman::p_load(tidyverse, ggplot2, ggstatsplot, plotly, scales, corrplot)
```

## 2. Loading the data

The processed fintech customer dataset used in the exploratory analysis is loaded again for confirmatory analysis.

```{r}
data_cleaned <- read_csv("data/colombia_customers_processed.csv")
```

## 3. Confirmatory Data Analysis

The following analysis evaluates key relationships between customer behavior indicators and churn probability identified during the exploratory data analysis stage. These analyses focus on customer sentiment, engagement, product adoption, and service interactions to assess their association with churn risk.

### 3.1 Customer Sentiment and Churn

Customer sentiment indicators such as Net Promoter Score (NPS) may influence customer retention behavior. To evaluate this relationship, churn probability is examined against NPS scores.

```{r}
p_nps <- ggplot(data_cleaned, aes(x = nps_score, y = churn_probability)) +
  geom_jitter(alpha = 0.15, width = 1, height = 0.01, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "darkred") +
  labs(
    title = "Churn Probability vs NPS Score",
    x = "NPS Score",
    y = "Churn Probability"
  ) +
  theme_minimal()

p_nps
```

Customers with higher NPS scores tend to exhibit lower churn probability. The visualization suggests a negative relationship between customer sentiment and churn risk, indicating that more satisfied or loyal customers are less likely to churn.

### 3.2 Satisfaction Score and Churn

Customer satisfaction is another important indicator of retention behavior. To examine this relationship, churn probability is compared across satisfaction score categories.

```{r}
ggplot(
  data_cleaned,
  aes(
    x = factor(satisfaction_score, ordered = TRUE),
    y = churn_probability,
    fill = factor(satisfaction_score, ordered = TRUE)
  )
) +
geom_boxplot(show.legend = FALSE) +
labs(
  title = "Churn Probability across Satisfaction Score Levels",
  x = "Satisfaction Score",
  y = "Churn Probability"
) +
theme_minimal()
```

Higher satisfaction score categories generally correspond to lower churn probability. Customers reporting stronger satisfaction with the platform tend to demonstrate more stable retention patterns compared with customers reporting lower satisfaction levels

### 3.3 Support Interactions and Churn

Customer support interactions may reflect service issues or dissatisfaction. To examine whether frequent support interactions are associated with churn, the average churn probability is calculated across support ticket counts.

```{r}
support_summary <- data_cleaned %>%
  group_by(support_tickets_count) %>%
  summarise(
    avg_churn = mean(churn_probability, na.rm = TRUE),
    n = n()
  )

ggplot(support_summary, aes(x = support_tickets_count, y = avg_churn)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "darkred", size = 2.5) +
  geom_text(aes(label = n), vjust = -0.8, size = 3) +
  labs(
    title = "Average Churn Probability by Number of Support Tickets",
    x = "Number of Support Tickets",
    y = "Average Churn Probability"
  ) +
  theme_minimal()
```

Churn probability shows some variation across different levels of support ticket activity. Moderate increases in support interactions appear associated with slightly higher churn risk, which may indicate that frequent service issues or support needs are linked to potential dissatisfaction.

### 3.4 Customer Tenure and Churn

Customer tenure may influence retention behavior, as long-term customers are often more integrated into the platform. This analysis examines the relationship between customer tenure and churn probability.

```{r}
ggplot(
  data_cleaned,
  aes(
    x = customer_tenure,
    y = churn_probability
  )
) +
geom_point(alpha = 0.08, color = "steelblue") +
geom_smooth(method = "loess", se = TRUE, color = "darkred") +
labs(
  title = "Churn Probability by Customer Tenure",
  x = "Customer Tenure",
  y = "Churn Probability"
) +
theme_minimal()
```

Customers with longer tenure generally exhibit lower churn probability. This suggests that customers who remain with the platform for extended periods become more established users and are less likely to leave.

### 3.5 Product Adoption and Churn

```{r}
ggplot(
  data_cleaned,
  aes(
    x = factor(active_products),
    y = churn_probability,
    fill = factor(active_products)
  )
) +
geom_boxplot(show.legend = FALSE) +
labs(
  title = "Churn Probability by Number of Active Products",
  x = "Number of Active Products",
  y = "Churn Probability"
) +
theme_minimal()
```

Customers holding a greater number of active products demonstrate noticeably lower churn probability. This pattern suggests that increased product adoption strengthens customer attachment to the platform and reduces the likelihood of churn.

## 4. CDA Summary

The confirmatory analysis examined several behavioral and sentiment indicators that may influence customer churn probability. The results suggest that customer satisfaction, product adoption, and platform engagement exhibit observable relationships with churn behavior.

Higher satisfaction and NPS scores are generally associated with lower churn probability, indicating that positive customer sentiment contributes to stronger retention. Similarly, customers with a greater number of active financial products tend to exhibit lower churn risk, suggesting that deeper integration with the platform increases customer loyalty.

Behavioral indicators such as customer tenure and support interactions also provide insight into churn dynamics. Longer-tenure customers appear to show more stable retention patterns, while increased support ticket activity may reflect potential service issues that contribute to higher churn risk.

Overall, these findings support the expectation that customer engagement, sentiment, and product adoption are important factors associated with churn probability within the fintech dataset.
