---
title: "Clustering Analysis"
author: "SU Bo-Han"
date: "March 13, 2026"
date-modified: "last-modified"
format: 
  html:
    theme: cosmo  
---

# **Load Packages and Data**

Check and load packages, ensuring they are sourced from and supported by CRAN.

```{r}
#| message: false
#| warning: false
pacman::p_load(tidyverse, scales, knitr,tidymodels,fpc)
df_raw <- read_csv("data/colombia_customers_processed.csv")
```

# **Data Processing**

-   We apply log10​ transformation to handle the long-tail distribution of transaction data, compressing outliers and normalizing the scale. This ensures the model captures behavioral trends more effectively without being distorted by extreme values.

-   To account for scale differences between value and volume, we normalize features to a mean of 0 and SD of 1. This eliminates unit bias, ensuring each metric carries equal weight in the clustering process.

```{r}
df_for_clustering <- df_raw %>%
  dplyr::select(customer_id, average_transaction_value, total_transaction_volume) %>%
  mutate(
    log_avg_value = log10(average_transaction_value + 1),
    log_total_vol = log10(total_transaction_volume + 1)
  )
cluster_recipe <- recipe(~ log_avg_value + log_total_vol, data = df_for_clustering) %>%
  step_normalize(all_predictors()) %>%
  prep()
df_cleaned <- bake(cluster_recipe, new_data = NULL)
summary(df_cleaned)
```

# **Feature Selection**

By selecting core transactional metrics, psychological traits (satisfaction, churn), and demographics, while applying log transformations to skewed values, we build a more precise and holistic customer profile. This streamlines model inputs for better performance while enhancing the ability to capture complex behavioral patterns.

```{r}
#| message: false
#| warning: false
df_for_analysis <- df_raw %>% 
  dplyr::select(
    customer_id, 
    total_transaction_volume, 
    average_transaction_value, 
    satisfaction_score, 
    churn_probability, 
    gender, 
    income_bracket
  ) %>% 
  mutate(
    log_total_vol = log10(total_transaction_volume + 1),
    log_avg_value = log10(average_transaction_value + 1)
  )
```

# **Data Validation & Cleaning**

```{r}
# ensure no missing values
colSums(is.na(df_for_analysis))
```

# **Representative Sampling**

By setting a random seed and sampling 5,000 records, we significantly enhance computational efficiency while maintaining data representativeness. This reduces processing time for visualization and modeling while ensuring that the analysis remains consistent and reproducible across different runs.

```{r}
set.seed(123) 
df_sampled <- df_for_analysis %>%
  sample_n(5000) 
```

# **Elbow Plot**

We iteratively evaluate k values from 1 to 10 and visualize the "Total Within-cluster Sum of Squares" through an Elbow Plot. This helps balance model complexity and precision, allowing us to mathematically identify the "elbow point" where improvements level off, ensuring we select the most meaningful number of clusters.

```{r}
#| message: false
#| warning: false
set.seed(123)
kclusts <- tibble(k = 1:10) %>%
  mutate(
    kmean = map(k, ~kmeans(df_cleaned, .x, nstart = 20)),
    glanced = map(kmean, glance)
  ) %>%
  unnest(glanced)

ggplot(kclusts, aes(k, tot.withinss)) +
  geom_line(size = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal() +
  labs(title = "Elbow Plot for K-means",
       x = "Number of Clusters (k)",
       y = "Total Within-cluster Sum of Squares")
```

# **Model Selection**

```{r}
#| include: false
df_cleaned_5000 <- bake(cluster_recipe, new_data = df_sampled)
cf_boot <- clusterboot(df_cleaned_5000, 
                       clustermethod = kmeansCBI, 
                       krange = 3, 
                       seed = 123)
```

```{r}
print(cf_boot$bootmean)
```

# **Clustering Analysis Dashboard**

We aggregate key metrics—transaction volume, value, and churn risk—to construct a behavioral profile for each segment. This visualization highlights the unique characteristics of each group, enabling us to translate abstract data clusters into actionable business personas and targeted strategies.

```{r}
set.seed(123)
final_kmeans <- kmeans(df_cleaned_5000, centers = 3, nstart = 25)
df_final_results <- df_sampled %>% 
  mutate(cluster = as.factor(final_kmeans$cluster))
df_final_results %>%
  group_by(cluster) %>%
  summarise(
    Across_Vol = mean(log_total_vol),
    Across_Avg = mean(log_avg_value),
    Across_Churn = mean(churn_probability, na.rm = TRUE)
  ) %>%
  pivot_longer(-cluster, names_to = "Metric", values_to = "Score") %>%
  ggplot(aes(x = Metric, y = Score, color = cluster, group = cluster)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Cluster Behavioral", subtitle = "Standardized metrics for cross-segment feature comparison")
```

```{r}
df_for_analysis <- df_raw %>% 
  dplyr::select(
    customer_id, 
    total_transaction_volume, 
    average_transaction_value, 
    satisfaction_score, 
    churn_probability, 
    occupation,     
    gender,         
    age_group      
  ) %>% 
  mutate(
    log_total_vol = log10(total_transaction_volume + 1),
    log_avg_value = log10(average_transaction_value + 1)
  )

set.seed(123)
df_sampled <- df_for_analysis %>% sample_n(min(5000, n()))

df_cleaned_5000 <- bake(cluster_recipe, new_data = df_sampled)
set.seed(123)
final_kmeans <- kmeans(df_cleaned_5000, centers = 3, nstart = 25)
df_final_results <- df_sampled %>%
  mutate(cluster = as.factor(final_kmeans$cluster))
p_occupation <- df_final_results %>%
  filter(!is.na(occupation)) %>% 
  mutate(occupation = fct_lump(occupation, n = 10)) %>% 
  ggplot(aes(x = cluster, fill = occupation)) + 
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::label_percent()) + 
  coord_flip() +
  labs(
    title = "Occupation Composition Analysis by Cluster", 
    x = "Cluster", 
    y = "Proportion", 
    fill = "Occupations"
  ) +
  theme_minimal()

print(p_occupation)
```

```{r}
df_final_results %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    avg_vol = mean(total_transaction_volume, na.rm = TRUE),
    avg_churn = mean(churn_probability, na.rm = TRUE),
    avg_sat = mean(satisfaction_score, na.rm = TRUE)
  )
```

```{r}
get_top_n_labels <- function(v, n = 3) {
  v <- na.omit(v)
  if (length(v) == 0) return("None")
  total_count <- length(v)
  top_n <- as.data.frame(table(v)) %>%
    rename(Occupation = v) %>%
    arrange(desc(Freq)) %>%
    slice_head(n = n) %>%
    mutate(label = paste0(Occupation, " (", round(Freq / total_count * 100, 1), "%)"))
  
  paste(top_n$label, collapse = ", ")
}
persona_summary_pro <- df_final_results %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    Top_3_Occupations = get_top_n_labels(occupation, n = 3),
    Avg_Volume = round(mean(total_transaction_volume, na.rm = TRUE), 0),
    Avg_Churn = round(mean(churn_probability, na.rm = TRUE), 2)
  )
print(persona_summary_pro)
```

-   **Segment 1 \| The High-Volume Core (**n=2,276): Comprising the largest portion of our sample, this group features stable professionals like **Bank Tellers and Dentists**. They represent our most reliable revenue stream with high transaction volumes.

-   **Segment 2 \| Premier High-Earners (**n=1,583): This elite group, dominated by **Accountants and Real Estate Agents**, yields the highest average transaction volume (\~59.7M). They are the prime targets for premium loyalty programs and high-tier financial products.

-   **Segment 3 \| Emerging Budget Segment (**n=1,141): Consisting of **Salespeople, Farmers, and Designers**, this group exhibits lower transaction volumes (\~14.6M). These customers may benefit from growth-oriented micro-services or entry-level financial tools.
