---
title: "Clustering Analysis"
author: "SU Bo-Han"
date: "March 13, 2026"
date-modified: "last-modified"
format: 
  html:
    theme: cosmo  
---

# **Load Packages and Data**

-   We begin by initializing the analytical environment with the `tidyverse` and `poLCA` ecosystems, ensuring a robust framework for data manipulation and latent class modeling.

```{r}
#| message: false
#| warning: false
pacman::p_load(tidyverse, poLCA, scales, patchwork, lubridate, knitr, dplyr)
df_raw <- read_csv("data/colombia_customers_processed.csv")
```

# **Latent Class Analysis (LCA)**

-   Feature selection is focused on two dimensions: **demographic traits** (e.g., income, occupation) and **behavioral indicators** (e.g., churn probability, satisfaction) to build a holistic segmentation framework.

```{r}
#| message: false
#| warning: false
df_sampled <- df_raw %>%
  dplyr::select(
    gender, income_bracket, occupation, age_group,        
    total_transaction_volume, churn_probability, satisfaction_score 
  )
```

# **Data Validation & Cleaning**

-   To maintain the integrity of our statistical inferences, we perform a rigorous quality check to ensure the dataset is free of missing values before proceeding to model estimation.

```{r}
# ensure no missing values
print(sapply(df_sampled, function(x) sum(is.na(x))))
```

# **Representative Sampling**

-   A random sample of 5,000 observations is extracted to achieve an optimal balance between computational efficiency and the statistical power required for representative results.

```{r}
set.seed(123) 
df_sampled <- df_sampled %>% sample_n(5000)
```

# **Binning & Factorization**

-   As LCA requires categorical inputs, continuous variables like transaction volume are discretized into ranked tiers, transforming raw metrics into meaningful behavioral levels.

```{r}
df_sampled <- df_sampled %>%
  mutate(
    vol_level = ntile(total_transaction_volume, 3), 
    churn_level = ntile(churn_probability, 3),
    sat_level = case_when(
      satisfaction_score <= 4 ~ 1,
      satisfaction_score <= 7 ~ 2,
      TRUE ~ 3
    )
  ) %>%
  dplyr::select(-total_transaction_volume, -churn_probability, -satisfaction_score)
df_lca_final <- df_sampled %>%
  mutate(across(everything(), factor)) %>%
  mutate(across(everything(), as.integer))
```

# **Determining Optimal K - BIC Plot**

-   Model selection is guided by the Bayesian Information Criterion (BIC). We iterate through multiple cluster counts to identify the "elbow" where model fit and parsimony are perfectly balanced.

```{r}
#| message: false
#| warning: false
f <- as.formula("cbind(gender, income_bracket, occupation, age_group, vol_level, churn_level, sat_level) ~ 1")

bic_list <- c()
for(i in 2:5){
  res <- poLCA(f, df_lca_final, nclass=i, nrep=2, maxiter=2000, verbose=FALSE)
  bic_list[i] <- res$bic
}
plot(2:5, bic_list[2:5], type="b", pch=19, col="#2E86C1", 
     main="Model Selection: BIC Plot", xlab="Number of Clusters", ylab="BIC Score")
```

# **Final Model Execution**

-   With the optimal K identified, we execute the final model to assign each customer to their most probable latent class, providing the basis for our segment profiles.

```{r}
set.seed(123)
lca_best <- poLCA(f, df_lca_final, nclass=3, nrep=10, maxiter=5000, verbose=FALSE)

df_results <- df_lca_final %>%
  mutate(cluster = as.factor(lca_best$predclass))
```

# **Clustering Analysis Dashboard**

-   This visualization translates complex model outputs into actionable insights, mapping the average behavioral signatures of each cluster to define their unique market characteristics.

```{r}
p1 <- df_results %>%
  group_by(cluster) %>%
  summarise(across(c(vol_level, churn_level, sat_level), mean)) %>%
  pivot_longer(-cluster, names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value, color = cluster, group = cluster)) +
  geom_line(size = 1.5, alpha = 0.8) + 
  geom_point(size = 4) +
  scale_y_continuous(limits = c(1, 3), breaks = 1:3, labels = c("Low", "Med", "High")) +
  labs(title = "Cluster Behavioral Profiles", 
       subtitle = "Line plots showing the average level of key metrics",
       x = "Behavioral Variable", y = "Average Level") +
  theme_minimal()
print(p1)
```

```{r}
lookup_table <- df_sampled %>%
  mutate(occ_id = as.integer(as.factor(occupation))) %>%
  dplyr::select(occupation, occ_id) %>%
  distinct()

p2_v2 <- df_results %>%

  left_join(lookup_table, by = c("occupation" = "occ_id")) %>% 
  ggplot(aes(x = cluster, fill = occupation.y)) + 
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent) +
  coord_flip() +
  labs(title = "Detailed Occupation Mix", 
       subtitle = "Who are these people? (By real job titles)",
       x = "Cluster", y = "Percentage Share", fill = "Job Title") +
  theme_minimal() +
  theme(legend.text = element_text(size = 8))

print(p2_v2)
```

```{r}
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
persona_summary <- df_results %>%
  group_by(cluster) %>%
  summarise(
    Segment_Size = n(),
    Top_Occupation = get_mode(occupation), 
    Top_Age_Group = get_mode(age_group),
    Top_Gender = get_mode(gender),
    Avg_Volume_Level = round(mean(as.integer(vol_level)), 1),
    Avg_Churn_Risk = round(mean(as.integer(churn_level)), 1),
    Avg_Satisfaction = round(mean(as.integer(sat_level)), 1)
  )

print(persona_summary)
```

```{r}
persona_table <- df_results %>%
  group_by(cluster) %>%
  summarise(
    Size = n(),
    Avg_Volume = round(mean(as.numeric(vol_level)), 2),
    Avg_Churn = round(mean(as.numeric(churn_level)), 2),
    Avg_Satisfaction = round(mean(as.numeric(sat_level)), 2)
  ) %>%
  mutate(Persona_Name = case_when(
    cluster == 1 ~ "Elite Loyalists",
    cluster == 2 ~ "At-Risk Dissatisfieds",
    cluster == 3 ~ "Stable Casual Users"
  )) %>%
  dplyr::select(cluster, Persona_Name, Size, everything())
kable(persona_table, 
      caption = "Final Customer Segmentation Summary (LCA Model)",
      align = "l")
```

-   **Cluster 1: Elite Loyalists** (Size: 1,344)

    -   **Profile:** High-skilled professionals (**Engineers, Doctors, Lawyers**).

    -   **Insight:** Despite being high-value, they have the **lowest satisfaction (1.16)**. They are "silent sufferers" who stay out of necessity but are unhappy with the service.

<!-- -->

-   **Cluster 2: At-Risk Dissatisfieds** (Size: 1,382)

    -   **Profile:** Service sector workers (**Sales, Waiters, Mechanics**).

    -   **Insight:** The most critical group with the **highest churn risk (2.86/3.0)** and low satisfaction. They are actively looking to leave the bank.

-   **Cluster 3: Stable Casual Users** (Size: 2,274)

    -   **Profile:** Stable middle-class roles (**Teachers, Accountants, Architects**).

    -   **Insight:** The most reliable segment. They exhibit the **lowest churn risk (1.45)** and the highest relative satisfaction, serving as the bank's stable revenue base.
